Strong lensing selection effects

### 2023-06-28 paper/seleff.tex
fixed refs and added Dogruel

At fixed minimum Einstein radius, there is a small difference between the median $\lmst$ in extended source and quasar lenses, typically of order 1\%.
The associated bias on $H_0$ is, therefore, equally small.
The value of $\lmst$, however, increases with increasing minimum $\tein$. This implies that if the two samples probe different ranges in Einstein radius, the bias on $H_0$ can be larger than that.
The bias can also be larger if the properties of the general population of galaxies (as opposed to the strong lenses) are used to inform the structure of time-delay lenses \citep[see][]{C+C16}. This is simply because quasar lenses are on average more similar to galaxy-galaxy lenses than they are to the underlying galaxy population (the blue and red curves in \Fref{fig:quasarbias} are closer to each other than they are to the dashed horizontal lines)
}

@@

{\bf
The simulations described in \Sref{sect:sims}, on which our experiment is based, are examples of so-called lens-selected surveys:
the lens samples are built by first defining a population of possible lens galaxies and then searching for strong lenses among them. Photometric lens searches are typically carried out in this way \citep[see e.g.][]{Gav++12,Son++18,Pet++19}.
An alternative approach is to build a source-selected sample, by first searching for signatures of strong lensing within a population of possible background sources and then obtaining information on the lens galaxies \citep[see e.g.][]{Hez++13}.
In principle, these two approaches lead to different strong lensing biases. %Therefore, strictly speaking, our results only apply to a lens-selected sample.
%For example, from \Fref{fig:lenspars} we can see that the distribution in redshift and stellar mass of the lenses extends beyond the bounds applied to the population of foreground galaxies: in principle, a source-based selection method 
For example, from \Fref{fig:lenspars} we can see that the stellar mass distribution of the lenses extends below the minimum mass cut that we applied to the population of foregrounds galaxies: a source-based search that is able detect those lenses would result in a different sample and a different associated strong lensing bias.
However, when considering only lenses with $\tein > 1''$, lens-selection and source-selection yield the same sample (see red histogram in \Fref{fig:lenspars}): in the limit of large Einstein radius, the two operations (identifying suitable lens galaxies and detecting a strongly lensed source) commute.

@@

We used the fundamental plane of early-type galaxies to set an upper limit on the intrinsic scatter parameters of the simulation: $\sigma_{\mathrm{sps}}$, $\sigma_h$ and $\sigma_\gamma$.
First, we measured the fundamental plane of the sample of SDSS early-type galaxies introduced in Appendix~\ref{sect:appendixa}.
We took measurements of the line-of-sight stellar velocity dispersion within the SDSS spectroscopic aperture, $\sigma_{\mathrm{ap}}$, from the SDSS data release 16 catalogue \citep{Ahu++20}.
Then, we fitted the following model to the distribution of $\sigma_{\mathrm{ap}}$ of the sample:
\begin{equation}\label{eq:fpfull}
\pr(\sigma_{\mathrm{ap}}) \sim \mathcal{N}(\mu_{\sigma} + \beta(\log{\mobs} - 11.4) + \xi(\log{\reff} - 1.2), \sigma_{\sigma}).
\end{equation}
We accounted for observational uncertainties on $\sigma_{\mathrm{ap}}$ when doing the fit, therefore the parameter $\sigma_{\sigma}$ describes the intrinsic scatter in the logarithm of the velocity dispersion, deconvolved from the observational scatter.
In principle we should also account for observational uncertainties on the stellar mass measurement, as they too contribute to the inferred scatter in $\sigma_{\mathrm{ap}}$. In practice, however, it is difficult to estimate observational uncertainties on stellar population synthesis measurements.
For this reason we chose not to propagate uncertainties on $\mobs$. As a result, the inferred scatter parameter $\sigma_{\sigma}$ is slightly overestimated.
We obtained $\mu_{\sigma}=2.36$, $\beta_{\sigma}=0.33$, $\xi_\sigma=-0.17$, and $\sigma_{\sigma}=0.035$.

>>>>

At fixed minimum Einstein radius, there is a small difference between the median $\lmst$ in extended source and quasar lenses, typically of order 1\%.
The associated bias on $H_0$ is, therefore, equally small.
The value of $\lmst$, however, increases with increasing minimum $\tein$. This implies that if the two samples probe different ranges in Einstein radius, the bias on $H_0$ can be larger than that.
The bias can also be larger if the properties of the general population of galaxies (as opposed to the strong lenses) are used to inform the structure of time-delay lenses \citep[see][]{C+C16}. This is simply because quasar lenses are on average more similar to galaxy-galaxy lenses than they are to the underlying galaxy population (the blue and red curves in \Fref{fig:quasarbias} are closer to each other than they are to the dashed horizontal lines).
}

@@

{\bf
The simulations described in \Sref{sect:lenspop}, on which our experiment is based, are examples of so-called lens-selected surveys:
the lens samples are built by first defining a population of possible lens galaxies and then searching for strong lenses among them. Photometric lens searches are typically carried out in this way \citep[see e.g.][]{Gav++12,Son++18a,Pet++19}.
An alternative approach is to build a source-selected sample, by first searching for signatures of strong lensing within a population of possible background sources and then obtaining information on the lens galaxies \citep[see e.g.][]{Hez++13}.
In principle, these two approaches lead to different strong lensing biases. %Therefore, strictly speaking, our results only apply to a lens-selected sample.
%For example, from \Fref{fig:lenspars} we can see that the distribution in redshift and stellar mass of the lenses extends beyond the bounds applied to the population of foreground galaxies: in principle, a source-based selection method 
For example, from \Fref{fig:lenspars} we can see that the stellar mass distribution of the lenses extends below the minimum mass cut that we applied to the population of foregrounds galaxies: a source-based search that is able detect those lenses would result in a different sample and a different associated strong lensing bias.
However, when considering only lenses with $\tein > 1''$, lens-selection and source-selection yield the same sample (see red histogram in \Fref{fig:lenspars}): in the limit of large Einstein radius, the two operations (identifying suitable lens galaxies and detecting a strongly lensed source) commute.

@@

We used the fundamental plane of early-type galaxies to set an upper limit on the intrinsic scatter parameters of the simulation: $\sigma_{\mathrm{sps}}$, $\sigma_h$ and $\sigma_\gamma$.
First, we measured the fundamental plane of the sample of SDSS early-type galaxies introduced in Appendix~\ref{sect:appendixa}.
We took measurements of the line-of-sight stellar velocity dispersion within the SDSS spectroscopic aperture, $\sigma_{\mathrm{ap}}$, from the SDSS data release 16 catalogue \citep{Ahu++20}.
Then, we fitted the following model to the distribution of $\sigma_{\mathrm{ap}}$ of the sample:
\begin{equation}\label{eq:fpfull}
\pr(\sigma_{\mathrm{ap}}) \sim \mathcal{N}(\mu_{\sigma} + \beta(\log{\mobs} - 11.4) + \xi(\log{\reff} - 1.2), \sigma_{\sigma}).
\end{equation}
We accounted for observational uncertainties on $\sigma_{\mathrm{ap}}$ when doing the fit, therefore the parameter $\sigma_{\sigma}$ describes the intrinsic scatter in the logarithm of the velocity dispersion, deconvolved from the observational scatter.
In principle we should also account for observational uncertainties on the stellar mass measurement, as they too contribute to the inferred scatter in $\sigma_{\mathrm{ap}}$. In practice, however, it is difficult to estimate observational uncertainties on stellar population synthesis measurements \citep[but see][]{Dog++23}.
For this reason we chose not to propagate uncertainties on $\mobs$. As a result, the inferred scatter parameter $\sigma_{\sigma}$ is slightly overestimated.
We obtained $\mu_{\sigma}=2.36$, $\beta_{\sigma}=0.33$, $\xi_\sigma=-0.17$, and $\sigma_{\sigma}=0.035$.
